Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-04 Paper/assa.tex
..

\subsection{Active subspaces}
\label{sub:ac}
As mentioned in the introduction, the active subspace is low-dimensional subspace
that constitutes important directions in a model's input
space~\cite{Constantine:2015}. The effective variability in the model output
due to uncertain inputs is predominantly captured along these directions.
Mathematically, these directions are the eigenvectors of a positive
semi-definite matrix, $\mat{C}$, defined as follows: (\alennote{Need to 
say what is the function $f$, first})
%
\be
\mat{C} = \int_\Omega (\nabla_{\bm{\xi}}f)(\nabla_{\bm{\xi}}f)^\top dP_{\bm{\xi}}
\label{eq:C}
\ee
%
where $\Omega$:~$[-1,1]^{N_p}$ is the domain of $\bm{\xi}$, and $N_p$ is the
number of uncertain inputs. \alennote{Why do we limit to $[-1, 1]^\Np$?} 
$dP_{\bm{\xi}}$ = $p(\bm{\xi})d\bm{\xi}$, where $p(\bm{\xi})$ is the joint probability
distribution of $\bm{\xi}$. Note that the function, $f$ is considered to be continuous 
and differentiable in the considered parameter domain. 
%Hence, it is possible that a given $f(\bm{\xi})$ might not admit an active
%subspace. However, it is of remarkable interest to investigate if one exists
%since the subspace could be exploited to reduce the dimensionality of the
%problem and hence the associated computational effort.
Here 
$\nabla_{\bm{\xi}}f$ denotes the gradient vector. 
%with individual components
%being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. 
Since $\mat{C}$ is symmetric and
positive semi-definite, it admits a spectral decomposition:
%
\be
\mat{C} = \bm{W}\bm{\Lambda}\bm{W}^\top
\ee
%
where $\bm{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with
$\lambda_i$'s sorted in descending order
\[
     \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_\Np \geq 0.
\] 
The eigenpairs are partitioned about the $i^{\text{th}}$ eigenvalue such that
the ratio, $\left(\frac{\lambda_i}{\lambda_{i+1}}\right)$ $\gg$ 1 as follows:
\be
 \bm{W} = [\bm{w}_1~\bm{w}_2],~~\bm{\Lambda} = \begin{bmatrix}\bm{\Lambda}_1 & \\  &
  \bm{\Lambda}_2 \\
\end{bmatrix}
\ee
 %
 where $\bm{w}_1$ comprises the dominant eigenspace of $\mat{C}$ regarded as
the active subspace, and $\bm{\Lambda}_1$ is the corresponding set of
eigenvalues. 
\alennote{$\vec{w}_1$ is a supposedly the matrix whose columns are 
the dominant eigenvectors. Denoting a matrix by a notation typically 
used for a vector is confusing. Use $\mat{W}_1$ instead?}

\alennote{The rest of this paragraph needs to be rewritten, as it is hard to 
see what is going on; what is $\theta$? I suggest, mentioning $\vec\theta$ 
and $\vec\xi$ earlier, and from then on make everything function of $\vec\xi$.
Explain the remaining ideas more clearly.}
The active subspace is used to construct the so-called active
variables, $\bm{w}_1^\top \bm{\xi}$ and $f(\bm{\xi}(\theta))$ is transformed as
$G(\bm{w}_1^\top \bm{\xi})$. In practice, the set of samples, $\bm{\xi}_i$'s
are projected to the physical space as $\bm{\theta}_i$'s at which the model
output, $f(\bm{\xi}(\theta))$ is computed.  The matrix, $\mat{C}$ is typically
approximated using pseudo-random sampling techniques such as Monte Carlo and
Latin hypercube sampling. The integral in~\eqref{eq:C} is replaced with a
summation as follows:
 %
 \be
 \mat{C}\approx \hat{\mat{C}} = \frac{1}{N}\sum\limits_{i=1}^{N} 
 (\nabla_{\bm{\xi}}f(\bm{\xi}_i))(\nabla_{\bm{\xi}}f(\bm{\xi}_i))^\top
 = \hat{\bm{W}}\hat{\bm{\Lambda}}\hat{\bm{W}}^\top
\label{eq:chat}
 \ee
 %
%%It has been shown that the accuracy of approximated dominant eigenspace,
%%$\hat{\bm{w}}_1$ is inversely proportional to the difference between the
%%smallest eigenvalue in $\hat{\bm{\Lambda}}_1$ and the largest eigenvalue in
%%$\hat{\bm{\Lambda}}_2$~\cite{Constantine:2014}. 
%%Components of the eigenvectors in the active subspace could be used for
%%estimating the so-called activity scores as a measure for global sensitivity
%%and also be used for approximating the DGSMs as discussed in the following
%%section.

>>>>

\subsection{Active subspaces}
\label{sub:ac}
As mentioned in the introduction, the active subspace is low-dimensional subspace
that constitutes important directions in a model's input
space~\cite{Constantine:2015}. The effective variability in the model output
due to uncertain inputs is predominantly captured along these directions.
Mathematically, for a given model output, $f(\bm{\xi})$, and a set of inputs, $\bm{\xi}$,
these directions are the eigenvectors of a positive
semi-definite matrix, $\mat{C}$, defined as follows: 
%
\be
\mat{C} = \int_\Omega (\nabla_{\bm{\xi}}f)(\nabla_{\bm{\xi}}f)^\top dP_{\bm{\xi}}
\label{eq:C}
\ee
%
where $\Omega\in\mathbb{R}^{N_p}$ is the domain of $\bm{\xi}$, and $N_p$ is the
number of uncertain inputs. 
$dP_{\bm{\xi}}$ = $p(\bm{\xi})d\bm{\xi}$, where $p(\bm{\xi})$ is the joint probability
distribution of $\bm{\xi}$. Note that the function, $f$ is considered to be continuous 
and differentiable in the considered parameter domain. 
%Hence, it is possible that a given $f(\bm{\xi})$ might not admit an active
%subspace. However, it is of remarkable interest to investigate if one exists
%since the subspace could be exploited to reduce the dimensionality of the
%problem and hence the associated computational effort.
Here 
$\nabla_{\bm{\xi}}f$ denotes the gradient vector. 
%with individual components
%being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. 
Since $\mat{C}$ is symmetric and
positive semi-definite, it admits a spectral decomposition:
%
\be
\mat{C} = \mat{W}\mat{\Lambda}\mat{W}^\top
\ee
%
where $\mat{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with
$\lambda_i$'s sorted in descending order
\[
     \lambda_1 \geq \lambda_2 \geq \cdots \geq \lambda_\Np \geq 0.
\] 
The eigenpairs are partitioned about the $i^{\text{th}}$ eigenvalue such that
the ratio, $\left(\frac{\lambda_i}{\lambda_{i+1}}\right)$ $\gg$ 1 as follows:
\be
 \mat{W} = [\mat{W}_1~\mat{W}_2],~~\mat{\Lambda} = \begin{bmatrix}\mat{\Lambda}_1 & \\  &
  \mat{\Lambda}_2 \\
\end{bmatrix}
\ee
 %
 where $\mat{W}_1$ comprises the dominant eigenspace of $\mat{C}$ regarded as
the active subspace, and $\mat{\Lambda}_1$ is the corresponding set of
eigenvalues.

\alennote{The rest of this paragraph needs to be rewritten, as it is hard to 
see what is going on; what is $\theta$? I suggest, mentioning $\vec\theta$ 
and $\vec\xi$ earlier, and from then on make everything function of $\vec\xi$.
Explain the remaining ideas more clearly.}
The active subspace is used to construct the so-called active
variables, $\mat{W}_1^\top \bm{\xi}$ and $f(\bm{\xi}(\theta))$ is transformed as
$G(\mat{W}_1^\top \bm{\xi})$. In practice, the set of samples, $\bm{\xi}_i$'s
are projected to the physical space as $\bm{\theta}_i$'s at which the model
output, $f(\bm{\xi}(\theta))$ is computed.  The matrix, $\mat{C}$ is typically
approximated using pseudo-random sampling techniques such as Monte Carlo and
Latin hypercube sampling. The integral in~\eqref{eq:C} is replaced with a
summation as follows:
 %
 \be
 \mat{C}\approx \hat{\mat{C}} = \frac{1}{N}\sum\limits_{i=1}^{N} 
 (\nabla_{\bm{\xi}}f(\bm{\xi}_i))(\nabla_{\bm{\xi}}f(\bm{\xi}_i))^\top
 = \hat{\mat{W}}\hat{\mat{\Lambda}}\hat{\mat{W}}^\top
\label{eq:chat}
 \ee
 %
%%It has been shown that the accuracy of approximated dominant eigenspace,
%%$\hat{\bm{w}}_1$ is inversely proportional to the difference between the
%%smallest eigenvalue in $\hat{\bm{\Lambda}}_1$ and the largest eigenvalue in
%%$\hat{\bm{\Lambda}}_2$~\cite{Constantine:2014}. 
%%Components of the eigenvectors in the active subspace could be used for
%%estimating the so-called activity scores as a measure for global sensitivity
%%and also be used for approximating the DGSMs as discussed in the following
%%section.

### 2018-08-04 Paper/conc.tex
..

%%Brief summary
%-Mathematical connections between Sobol, DGSMs, and activity scores exist and could
%be exploited to perform GSA with reduced effort.
%-Iterative implementation yields advantage
%-Consistency between the two approaches with DGSM as well as among themselves
%encourages the use of grad-free.
%-Grad-free was found to be suitable for a high-dimensional problem.
%-Based on the results presented, the following parameter were found to be 
%significant.
%
%discussion
%- Methodology is agnostic to the choice of mechanism and could be applied to other
%systems.
%- However, constraints exist in terms of differentiability of the QoI. Not all
%input/output relationships exhibit an active subspace. However, it is important to
%check.
%- Activity scores could be used to identify key reactions and thereby construct
%reduced order mechanisms for reducing computational effort and investigating the
%inadequacy associated with rate laws.

@@

Two computational strategies were explored to compute the active subspace for the H$_2$/O$_2$
kinetics application in this work. The first strategy referred to as grad-based requires computation of gradient of the 
model output with respect to the input, whereas, the second strategy, referred to as grad-free implements a local
linear approximation to the model output in order to estimate the gradient. The two strategies are shown to yield
consistent results for the 19-dimensional problem whereby only the pre-exponents are considered to be uncertain.
Additionally, the activity scores are also shown to be consistent with the screening metric estimates based on
DGSMs in~\cite{Vohra:2018}. Note that an iterative procedure is adopted for implementing the two strategies
for enhancing computational efficiency. 
Motivated by these findings, the grad-free approach is used for computing the active subspace in the high-dimensional
case involving 33 uncertain parameters. Based on the prior marginals of the uncertain parameters, our results
indicate that the uncertainty in the ignition delay is largely dependent on $A_1$, $A_9$, and $E_{a,1}$, although
 contributions from $A_{15}$ and $A_{17}$ are also found to be significant.

>>>>

%%Brief summary
%-Mathematical connections between Sobol, DGSMs, and activity scores exist and could
%be exploited to perform GSA with reduced effort.
%-Iterative implementation yields advantage
%-Consistency between the two approaches with DGSM as well as among themselves
%encourages the use of gradient-free.
%-Grad-free was found to be suitable for a high-dimensional problem.
%-Based on the results presented, the following parameter were found to be 
%significant.
%
%discussion
%- Methodology is agnostic to the choice of mechanism and could be applied to other
%systems.
%- However, constraints exist in terms of differentiability of the QoI. Not all
%input/output relationships exhibit an active subspace. However, it is important to
%check.
%- Activity scores could be used to identify key reactions and thereby construct
%reduced order mechanisms for reducing computational effort and investigating the
%inadequacy associated with rate laws.

@@

Two computational strategies were explored to compute the active subspace for the H$_2$/O$_2$
kinetics application in this work. The first strategy referred to as gradient-based requires computation of gradient of the 
model output with respect to the input, whereas, the second strategy, referred to as gradient-free implements a local
linear approximation to the model output in order to estimate the gradient. The two strategies are shown to yield
consistent results for the 19-dimensional problem whereby only the pre-exponents are considered to be uncertain.
Additionally, the activity scores are also shown to be consistent with the screening metric estimates based on
DGSMs in~\cite{Vohra:2018}. Note that an iterative procedure is adopted for implementing the two strategies
for enhancing computational efficiency. 
Motivated by these findings, the gradient-free approach is used for computing the active subspace in the high-dimensional
case involving 33 uncertain parameters. Based on the prior marginals of the uncertain parameters, our results
indicate that the uncertainty in the ignition delay is largely dependent on $A_1$, $A_9$, and $E_{a,1}$, although
 contributions from $A_{15}$ and $A_{17}$ are also found to be significant.

### 2018-08-04 Paper/intro.tex
..

We are interested in investigating the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on ignition delay. The $A_i$'s are considered to be
uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem computationally challenging. 
To reduce the effort, we explore two
strategies: 
a gradient-based approach and  
a gradient-free approach.
%involving local linear approximation which does notrequire gradient computation
%(grad-free).  
\alennote{Thorough the paper 
change grad-based to gradient-based, and similarly for grad-free}. 
%
These strategies are based on algorithms 1.1 and 1.2 
in~\cite{Constantine:2015}. In this work, an
iterative approach is adopted for implementing the two strategies.
Consequently, the amount of computational effort appears to have decreased
\alennote{This will be unclear to readers}, and
is found to be relatively less in the case of grad-free approach as discussed
later in section~\ref{sec:method}.

\alennote{This paragraph will not be meaningful to readers, especially those not familiar
with active subspaces; Need to explain DGSMs, their importance, 
and say something general about our
approach for approximating them.}
Components of the dominant eigenvectors can be used to estimate the so-called
activity scores~\cite{Diaz:2016,Constantine:2017} which provide an insight into
the relative importance of the uncertain parameters. In section~\ref{sub:gsa},
we demonstrate that the activity scores could be used to approximate the
derivative-based global sensitivity measures (DGSMs)~\cite{Sobol:2009}, and
establish a relationship between the Sobol' total effect index, the DGSMs, and
the activity scores.

%Below, we summarize key contributions of the paper and its
%organization.

\paragraph{Contributions}
Main contributions of this paper are as follows: (1) A perspective on activity
scores as means for approximating the DGSMs is provided. (2) A general
relationship between DGSMs, activity scores, and the total Sobol' index is
established. (3) Iterative gradient-based and gradient-free approaches for
computing the active subspace are demonstrated. The two approaches are shown to
be consistent and yield greater computational gains compared to the
conventional approach. As mentioned above, the gains are greater in the case of
grad-free approach. (4) The grad-free approach is implemented for discovering
an active subspace in a 33-dimensional parameter space for the H$_2$/O$_2$
reaction kinetics application. 
\alennote{This says too much. Focus on the actual novel 
contributions; also, the application should be made more prominent.}

 \paragraph{Organization}
 The paper is organized as follows. In Section~\ref{sub:ac}, a brief theoretical background on
 the active subspace methodology is provided. In Section~\ref{sub:gsa}, it is shown that the activity
 scores provide a reasonable approximation to the DGSMs especially in a high-dimensional setting. 
 Additionally, a relationship between the three global sensitivity measures, namely, the activity scores,
 DGSMs, and the total Sobol' indices is established. In Section~\ref{sec:method}, a systematic framework
 for computing the active subspace using the grad-based, and the grad-free approaches is presented. 
The two approaches are implemented to the H$_2$/O$_2$ reaction kinetics problem whereby only
the $A_i$'s are considered as uncertain. Our findings illustrate the consistency in the results based on
the two approaches thereby encouraging the use of the grad-free approach. In Section~\ref{sec:app},
we further explore the suitability of the grad-free approach by implementing it to a higher-dimensional
H$_2$/O$_2$ reaction kinetics problem by accounting for the uncertainty in $E_{a,i}$'s as well. 
Finally, we draw important conclusions based on the various aspects of this work in Section~\ref{sec:conc}.

>>>>

We are interested in investigating the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on ignition delay. The $A_i$'s are considered to be
uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem high-dimensional and computationally challenging. 
For the purpose of dimension reduction, the active subspace is evaluated using
two computational strategies: a gradient-based approach and  
a gradient-free approach. These strategies are based on algorithms 1.1 and 1.2 
in~\cite{Constantine:2015}. As mentioned earlier, the gradient information of the model
output is needed in both cases. However, in the case of gradient-free approach,
the gradient is approximated using a linear regression-fit to the model output.
Hence, the amount of computational effort is reduced in this case. Both strategies
are implemented in an iterative manner i.e. a new set of model evaluations are
generated at each iteration until convergence is accomplished as discussed
later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global sensitivity
measures associated with uncertain inputs of a model. Uncertainty along the inputs with
a relatively small estimate of the sensitivity measure can be ignored for the purpose of
reducing the computational effort associated with uncertainty quantification (UQ). The
Sobol' indices are commonly used for this purpose~\cite{Sobol:2001}. However, obtaining
converged estimates of the Sobol' indices typically requires tens of thousands of model
runs and is therefore impractical in situations involving intensive model simulations.
Recently, the derivative-based global sensitivity measures~(DGSMs) have been 
developed to approximate the upper bound on the Sobol' indies with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. In Section~\ref{sec:assa}, we show
that the DGSMs can in fact be approximated using the components of the eigenvectors
in the active subspace.

Main contributions of this paper are as follows: (1) Mathematical linkages between the
global sensitivity measures: Total Sobol index, DGSMs, and the activity scores are
established; (2) An efficient gradient-free iterative approach based on local linear
approximation of the model output is shown to yield consistent results with the
gradient-based approach. The two approaches were implemented to compute
the active subspace for a 19-dimensional H$_2$/O$_2$ kinetics problem with 
uncertain pre-exponents in the rate law for individual reactions. (3) The gradient-free
approach is finally implemented to discover an active subspace for the 33-dimensional
H$_2$/O$_2$ reaction kinetics application in which case the pre-exponents as well
as the activation energies are considered to be uncertain.

 The remaining of this paper is organized as follows. In Section~\ref{sub:ac}, a brief theoretical background on
 the active subspace methodology is provided. In Section~\ref{sub:gsa}, it is shown that the activity
 scores provide a reasonable approximation to the DGSMs especially in a high-dimensional setting. 
 Additionally, a relationship between the three global sensitivity measures, namely, the activity scores,
 DGSMs, and the total Sobol' indices is established. In Section~\ref{sec:method}, a systematic framework
 for computing the active subspace using the gradient-based, and the gradient-free approaches is presented. 
The two approaches are implemented to the H$_2$/O$_2$ reaction kinetics problem whereby only
the $A_i$'s are considered as uncertain. Our findings illustrate the consistency in the results based on
the two approaches thereby encouraging the use of the gradient-free approach. In Section~\ref{sec:app},
the gradient-free approach is implemented to the 33-dimensional
H$_2$/O$_2$ reaction kinetics application by accounting for the uncertainty in $E_{a,i}$'s as well. The
resulting active subspace is found to be 1-dimensional indicating enormous potential for reduction in
computational effort pertaining to UQ, in this case. 
Finally, we draw important conclusions based on the various aspects of this work in Section~\ref{sec:conc}.
